%%% Local Variables:
%%% TeX-master: "Proofs"
%%% End:
\probsec{~\ref{sec:conjunctions}}
\begin{enumerate}
    \item You meet Allen, Burroughs, and Corso.
  \begin{dialogue}
    \speak{Allen} Burroughs and Corso are \knights.
    \speak{Burroughs} Either I or Allen is a \knave.
  \end{dialogue}
  What can you conclude?

    \item You meet Anand, Botvinnik, and Capablanca.
  \begin{dialogue}
    \speak{Anand} Either Botvinnik is a \knave or Capablanca is a \knight.
    \speak{Botvinnik} Capablanca and I are both \knaves.
  \end{dialogue}
  What can you conclude?

    \item You encounter Asia, Brussels, and Cuba.
  \begin{dialogue}
    \speak{Asia} Cuba and I are both \knights.
    \speak{Brussels} Either Asia or Cuba is a \knave.
    \speak{Cuba} Asia and Brussels are both \knaves.
  \end{dialogue}
  What can you conclude?

    \item You encounter Al, Bonnie, Clyde, and Dillinger. One of them is a \knave who stole a very valuable mathematical manuscript. Determine from their statements who is the thief.
  \begin{dialogue}
    \speak{Al} Dillinger is the thief.
    \speak{Bonnie} At least one of us is a \knight.
    \speak{Clyde} At least one of us is a \knave.
    \speak{Dillinger} Both Bonnie and Clyde are \knights.
  \end{dialogue}

    \item You meet Arkana, Bullard, Robin, and Corinthian.
  \begin{dialogue}
    \speak{Arkana} Either Bullard is a \knight or Robin is a \knave.
    \speak{Bullard} Either Arkana or Corinthian is a \knave.
    \speak{Robin} At least one of the other three is a \knave.
  \end{dialogue}
  What can you conclude?

    \item You run into Agamemnon, Briseis, Cassandra, and Diomedes.
  \begin{dialogue}
    \speak{Agamemnon} Briseis is a \knight and Cassandra is a \knave.
    \speak{Briseis} Cassandra is a \knave and Diomedes is a \knave.
    \speak{Cassandra} Either Agamemnon or Briseis is a \knave.
    \speak{Diomedes} Cassandra is a \knave or Agamemnon is a \knight.
  \end{dialogue}
  What can you conclude?


\end{enumerate}
